Inferring the gravitational potential of the Milky Way with a few precisely measured stars

### 2014-02-08 text/paper2/ms.tex
model section

@@

Take $\D$ to be the observed, heliocentric 6D position of a star --- e.g., the measured position on the sky, ($l$, $b$); distance, $D$; proper motions, ($\mu_l$, $\mu_b$); and line-of-sight velocity, $v_r$ --- and take $\W$ to be the corresponding \emph{true}, 6D position of the star in the same coordinate and reference frame. Assume that we have determined that this star was once part of a progenitor system (e.g., satellite galaxy) tidally disrupting and forming a cold debris structure in the potential, $\Phi$, of some parent galaxy. The position of the satellite is observed to be at heliocentric 6D position $\D_s$, with true position $\W_s$ (where any subscript $s$ refers to the satellite). The true 6D spherical, heliocentric positions of the star and satellite -- $\W$ and $\W_s$ -- have corresponding 6D cartesian, Galactocentric coordinate vectors $\X$ and $\X_s$. Initially we assume the present-day mass of the satellite, $M_s$, and velocity dispersion, $\sigma_v$, have been measured perfectly and neglect mass-loss from the satellite. With this notation, we write the joint posterior probability for the observed position of a star, satellite, and parameters for the parent potential as:
\begin{align}
	p(\Phi, \W, \W_s, \tau \given \D, \D_s) &= \frac{1}{\mathcal{Z}} p(\D \given \W) p(\D_s \given \W_s) 
												       p(\W \given \W_s, \tau, \Phi) 
												       p(\Phi) p(\tau)
\end{align}
where the factor $\mathcal{Z}$ only depends on the properties of the data (the \emph{evidence}), and is thus constant when varying the model parameters. We assume that the observational errors are Gaussian in heliocentric coordinates:
\begin{align}
	p(\D \given \W) &= \mathcal{N}(\W \given \D, \bSigma)\\
	p(\D_s \given \W_s) &= \mathcal{N}(\W_s \given \D_s, \bSigma_s)
\end{align}
where the covariance matrices $\bSigma$ and $\bSigma_s$ specify the observational variances on the observed 6D position of star and the satellite, respectively.

The disruption of a satellite galaxy due to tidal shocking [...] Thus, we model the present-day position and velocity of the disrupted star as the result of sampling from isotropic, log-normal distributions centered on the instantaneous tidal radius, $R_{\rm tide}$, and velocity dispersion, $\sigma_v$, of the satellite at some time $\tau$ in the past, evolved to present day in the parent potential, $\Phi$. In this model, we neglect the effect of self-interaction with the halo of the progenitor [TODO: maybe we need to justify this, add APW's figure from paper 1 ref. response? refer to another section]. The likelihood of a star's position today is computed by first transforming the positions to Galactocentric, cartesian coordinates and integrating the orbits of the satellite and star backwards for several Gigayears in the parent potential, $\Phi$, treating both as test particles. The distance, $R$, and velocity, $V$, of the star relative to the satellite are computed at each time-step, and the likelihood is computed as the product of an isotropic, log-normal centered on the tidal radius with an isotropic, log-normal centered on the velocity dispersion of the satellite:
\begin{align}
	p(\W \given \W_s, \tau, \Phi) &= p(\X \given \X_s, \tau, \Phi) \jac{(l,b,D,\mu_l,\mu_b,v_r)}\\
	p(\X \given \X_s, \tau, \Phi) &= p(\ln R, \ln V \given \tau, \Phi) \jac{(x,y,z,v_x,v_y,v_z)}\label{eq:int} \\
	p(\ln R, \ln V \given \tau, \Phi) &= \mathcal{N}(\ln R(\tau) | \ln \rtide(\tau))\mathcal{N}(\ln V(\tau) | \ln \sigma_v(\tau))
\end{align}
where in eq.~\ref{eq:int} we have implicitly integrated over and neglected the two angle coordinates associated with $\ln R$ and the two angles associated with $\ln V$.

% , but the vectors $\D$ and $\W$ [exist] in different coordinate systems --- heliocentric spherical and Galactocentric cartesian, respectively --- so we must also track the Jacobian that [describes] this transformation:

In general, $p(\tau)$ is likely conditional on the observed orbit of the progenitor---stars are preferentially stripped at pericentric passages \citep{??}---but this is neglected and assumed to be uniform over the entire interaction history between the satellite and parent. Similarly, the priors on potential parameters, $p(\Phi)$, are taken to be uniform.

>>>>

\newcommand{\bsr}{\boldsymbol r}
\newcommand{\bsv}{\boldsymbol v}

@@

[Conceptual description of the model]

Take $\D$ to be the observed, heliocentric 6D position of a star --- e.g., the measured position on the sky, ($l$, $b$); distance, $D$; proper motions, ($\mu_l$, $\mu_b$); and line-of-sight velocity, $v_r$ --- and take $\W$ to be the corresponding \emph{true}, 6D position of the star in the same coordinate and reference frame. Assume that we have determined that this star was once part of a progenitor system (e.g., satellite galaxy) tidally disrupting and forming a cold debris structure in the potential, $\Phi$, of some parent galaxy. The position of the satellite is observed to be at heliocentric 6D position $\D_s$, with true position $\W_s$ (where any subscript $s$ refers to the satellite). The true 6D heliocentric spherical positions of the star and satellite -- $\W$ and $\W_s$ -- have corresponding 6D Galactocentric cartesian coordinate vectors $\X$ and $\X_s$ comprised of position and velocity vectors ($\bsr$,$\bsv$) and ($\bsr_s$,$\bsv_s$). Initially we assume the present-day mass of the satellite, $M_s$, and velocity dispersion, $\sigma_v$, have been measured perfectly and neglect mass-loss from the satellite. With this notation, we write the joint posterior probability for the observed position of a star, satellite, and parameters for the parent potential as:
\begin{align}
	p(\Phi, \W, \W_s, \tau \given \D, \D_s) &= \frac{1}{\mathcal{Z}} p(\D \given \W) p(\D_s \given \W_s) 
												       p(\W \given \W_s, \tau, \Phi) 
												       p(\Phi) p(\tau)
\end{align}
where the factor $\mathcal{Z}$ only depends on the properties of the data (the \emph{evidence}), and is thus constant when varying the model parameters. We assume that the observational errors are Gaussian in heliocentric coordinates:
\begin{align}
	p(\D \given \W) &= \mathcal{N}(\W \given \D, \bSigma)\\
	p(\D_s \given \W_s) &= \mathcal{N}(\W_s \given \D_s, \bSigma_s)
\end{align}
where the covariance matrices $\bSigma$ and $\bSigma_s$ specify the observational variances on the observed 6D position of star and the satellite, respectively.

The disruption of a satellite galaxy due to tidal shocking [...connect to Choi work some how?] We model the present-day position and velocity of a disrupted star as the result of sampling from isotropic, log-normal distributions in relative distance, $R=\left\vert \bsr-\bsr_s \right\vert$, and relative velocity, $V=\left\vert \bsv-\bsv_s \right\vert$, at the time the star comes unbound from the satellite, evolved to present day in the parent potential, $\Phi$. The distributions over $\ln R$ and $\ln V$ are centered on the logarithm of the instantaneous tidal radius, $R_{\rm tide}$, and velocity dispersion, $\sigma_v$, of the satellite at time $\tau$: $\mathcal{N}(\ln R \given \ln\rtide)\vert_{t=\tau}$ and $\mathcal{N}(\ln V \given \ln \sigma_v)\vert_{t=\tau}$. In this model, we neglect the effect of self-interaction with the halo of the progenitor [TODO: maybe we need to justify this, add APW's figure from paper 1 ref. response? refer to another section].

The likelihood of a star's position today is computed by first transforming the observed position to Galactocentric cartesian coordinates and integrating the orbits of the satellite and star backwards for several Gigayears in the parent potential, $\Phi$, treating both as test particles. The distance, $R$, and velocity, $V$, of the star relative to the satellite are computed at each time-step, and the likelihood is computed as the product of the isotropic, log-normal centered on the tidal radius with the isotropic, log-normal centered on the velocity dispersion of the satellite:
\begin{align}
	p(\W \given \W_s, \tau, \Phi) &= p(\X \given \X_s, \tau, \Phi) \jac{(l,b,D,\mu_l,\mu_b,v_r)}\\
	p(\X \given \X_s, \tau, \Phi) &= p(\ln R, \ln V \given \tau, \Phi) \jac{(x,y,z,v_x,v_y,v_z)}\label{eq:int} \\
	p(\ln R, \ln V \given \tau, \Phi) &= \mathcal{N}(\ln R | \ln \rtide)\mathcal{N}(\ln V | \ln \sigma_v)\vert_{t=\tau}
\end{align}
% , but the vectors $\D$ and $\W$ [exist] in different coordinate systems --- heliocentric spherical and Galactocentric cartesian, respectively --- so we must also track the Jacobian that [describes] this transformation:
where in eq.~\ref{eq:int} we have implicitly integrated over and neglected the two angle coordinates associated with $\ln R$ and the two angles associated with $\ln V$. We then compute the marginal likelihood, $p(\W \given \W_s, \Phi)$, by integrating over all possible unbinding times,
\begin{align}
	p(\W \given \W_s, \Phi) = \int p(\W \given \W_s, \tau, \Phi)p(\tau) d\tau.
\end{align}
In general, $p(\tau)$ is likely conditional on the orbit of the progenitor --- stars are preferentially stripped at pericentric passages --- but this is neglected and is taken to be uniform over the entire interaction history between the satellite and parent.
